Efficient forwards-time simulation with ancestral recombination graphs

### 2017-10-20 forwards_paper.tex

In so doing we record the \emph{population pedigree} --
the entire history of parent-offspring relationships of an entire population going back to a remote time --
as well as information encoding the genetic outcomes of each ancestral meiosis --
who inherited which parts of which parental chromosomes.
This embellished graph is known as the \emph{ancestral recombination graph}, or {ARG} \citep{griffiths}.
Combined with ancestral genotypes and the origins of new mutations,
it completely specifies the genomic sequence of any individual in the population at any time.
However, much less than the entire ARG is needed to specify relationships
between any given set of samples --
only those portions of it from which those samples have actually inherited,
back to their most recent common ancestors.
The assumptions of coalescent theory
% -- random mating and neutrality --
imply that this random set describes a Markov process looked at \emph{backwards} in time,
and can hence be simulated without reference to the unnecessary remainder
\citet{griffithsmarjoram,msprime}.
This stochastic process is also called the ``ARG'',
but we reserve this name for the embellished graph itself.
% Although this point was in principle known since \citet{griffithsmarjoram},
% only recent algorithmic advances made it possible to actually simulate the process correctly,
% without approximation, across whole chromosomes \citep{msprime}.

@@

%%%%%%
\subsection*{Data structures:}

First we describe the data structures we use for recording genealogical history,
as implemented in \msprime{}.
These derive from those described in \citet{msprime},
but have been modified to remove redundancy.
The tables below give the example tree sequence of Figure \ref{fig:ex_tree_seq}.

@@

To clarify terminology,
below a \emph{tree} refers to
a genealogical tree describing how a collection of individuals are related to each other.
A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
tree relating all samples to each other at any point along the genome.
In the context of a tree sequence,
\emph{nodes} refer to distinct ancestors,
and may be identified with points in the trees of a tree sequence.
Every tip or branching point in a tree is associated with a node.
A tree sequence does not necessarily carry complete genealogical information about every node,
but calling a node a \emph{sample} implies that complete information is present for that ancestor,
i.e., the node is associated with a point in every tree in the tree sequence.
Since each node represents a certain ancestor, it has a
unique ``time'', thought of as her birth time, which determines the height
of any branching points she is associated with.  A given node will be
associated with branching points of all trees across a region if that node
is the most recent common ancestor to the subtending tips across that region.
This information is stored in the columns of a \textbf{Node Table}:
\begin{center}
\begin{tabular}{cccc}
    id   &  is\_sample &  population &  time \\
    0    &  1          & 0           & 0 \\
    1    &  1          & 0           & 0 \\
    2    &  1          & 0           & 0 \\
    3    &  0          & 0           & 0.3 \\
    4    &  0          & 0           & 0.5 \\
    5    &  0          & 0           & 0.6 \\
    6    &  0          & 0           & 0.7 \\
\end{tabular}
\end{center}
where ``flags'' records other information
(e.g., a binary mask of `1' indicates the node is a sample).
Importantly, the \textbf{node ID} of a node is given implicitly
by the (zero-based) index of its corresponding row in the Node Table.

>>>>

In so doing we record the \emph{population pedigree} --
the entire history of parent-offspring relationships of an entire population going back to a remote time --
as well as information encoding the genetic outcomes of each ancestral meiosis --
who inherited which parts of which parental chromosomes.
This embellished graph is known as the \emph{ancestral recombination graph}, or {ARG} \citep{griffiths}.
Combined with ancestral genotypes and the origins of new mutations,
it completely specifies the genomic sequence of any individual in the population at any time.
However, much less than the entire ARG is needed to specify relationships
between any given set of samples --
only those portions of it from which those samples have actually inherited,
back to their most recent common ancestors.
The assumptions of coalescent theory
% -- random mating and neutrality --
imply that this random set describes a Markov process looked at \emph{backwards} in time,
and can hence be simulated without reference to the unnecessary remainder
\citet{griffithsmarjoram,msprime}.
This stochastic process is also called the ``ARG'',
but we reserve this name for the embellished graph itself.
% Although this point was in principle known since \citet{griffithsmarjoram},
% only recent algorithmic advances made it possible to actually simulate the process correctly,
% without approximation, across whole chromosomes \citep{msprime}.
\jk{I find this confusing I'm afraid. I agree we should mention the ARG but I think only in passing.
I have never found a situation in which an ARG made things easier to understand!}

@@

%%%%%%
\subsection*{Tree sequences}
A tree sequence is an efficient encoding for a sequence of correlated trees such as
is produced by recombination. The encoding is efficient because branches that are
shared by adjacent trees are stored once, rather than repeatedely for each tree.
The topology of a tree sequence is defined via a set of \emph{nodes} and
\emph{edges}. A node in a tree sequence refers to a distinct ancestor and
corresponds to the vertices in the individual genealogies along the sequence.
Since each node represents a specific ancestor, it has a unique ``time'',
thought of as her birth time, which determines the height of any vertices
she is associated with. In the example of Figure~\ref{fig:example_tree_sequence}
we have a total of five nodes. Nodes 0, 1 and 2 occur at time 0 and are our
samples. Nodes 3 and 4 represent the ancestors of these samples, and were born at
time 1.5 and 2.5 respectively.

\begin{figure}
    \begin{center}
        \includegraphics[width=0.6\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10. In the
        left-most panel we show the tree sequence pictorially in two different ways:
        in the top are the tree topologies and the bottom shows the spatial extent of the
        edges that define these topologies. The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations) defined
        by \msprime.
        \label{fig:example_tree_sequence}
    }
\end{figure}

The topology of a tree sequence is defined by the \emph{edges} which define how
nodes relate to each other over specific genomic intervals. Each edge is a tuple
$(\ell, r, p, c)$, where $[\ell, r)$ is a half-open genomic interval defining the
spatial extend of the edge; $p$ and $c$ are the parent and child nodes specifying
a single branch in the tree over this interval. The spatial extent of the
edges defining the topology of Figure~\ref{fig:example_tree_sequence} in the
bottom left panel. For example, the branch joining nodes 1 to 3 is shared in both trees,
and is recorded in a single edge extending over the whole chromosome. It is this
method of capturing the shared structure between adjacent trees that makes the
tree sequence encoding very compact and algorithmically efficient.

\jk{Continue along these lines with sites and mutations after updating the figure}

\jk{Other stuff we can review here:}
\begin{enumerate}
\item Encoding a tree sequence topology like this is very efficient, much, much more compact than newick.
Can refer to example in msprime paper, or could do a large running example here. Say 500K samples of
human CHR1, something like that. Newick will be hundreds of TB.

\item Can also make the point that it's much more efficient than, say, VCF. We could drop some
mutations on the example above, and then report how big it would be if encoded with VCF and BCF.

\item The \msprime\ API is very efficient for working with these large datasets. Examples on
how long it takes to iterate over all the trees, compute pi, etc.

\item The point of all this is to emphasise that not only is it efficient to generate the
data as a tree sequence, it's also very efficient to store and process it this way too. By the
time we get to the end of this section we should have convinced the reader of this.

\end{enumerate}

\subsection*{Data structures:}

First we describe the data structures we use for recording genealogical history,
as implemented in \msprime.
These derive from those described by \citet{kelleher2016efficient},
but have been generalised and modified to remove redundancy.
The tables below give the example tree sequence of Figure \ref{fig:ex_tree_seq}.

@@

To clarify terminology,
below a \emph{tree} refers to
a genealogical tree describing how a collection of individuals are related to each other.
A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
tree relating all samples to each other at any point along the genome.
In the context of a tree sequence, \emph{nodes} refer to distinct ancestors,
and correspond to the vertices in the trees of a tree sequence.
Since each node represents a certain ancestor, it has a unique ``time'', thought of as her birth time,
which determines the height of any branching points she is associated with.  A given node will be
associated with branching points of all trees across a region if that node
is the most recent common ancestor to the subtending tips across that region.
This information is stored in the columns of a \textbf{Node Table}:
\begin{center}
\begin{tabular}{cccc}
    id   &  is\_sample &  population &  time \\
    0    &  1          & 0           & 0 \\
    1    &  1          & 0           & 0 \\
    2    &  1          & 0           & 0 \\
    3    &  0          & 0           & 0.3 \\
    4    &  0          & 0           & 0.5 \\
    5    &  0          & 0           & 0.6 \\
    6    &  0          & 0           & 0.7 \\
\end{tabular}
\end{center}
where ``flags'' records other information
(e.g., a binary mask of `1' indicates the node is a sample).
Importantly, the \textbf{node ID} of a node is given implicitly
by the (zero-based) index of its corresponding row in the Node Table.
